---
title: 'Exercise: R Studio Tour'
output: html_document
date: "2024-09-03"
---

```{r setup, include=FALSE}
library(klippy)
knitr::opts_chunk$set(echo = TRUE)
klippy::klippy()
```

# **Objectives**

Get familiar with RStudio by working with an R script and reviewing the different panels in the IDE.

# **Exercise**

1.  Open RStudio

    a.  If in RStudio Cloud, click on the “New Project” Button.

    b.  If in RStudio on your desktop, open RStudio.

2.  Create a new R script `File> New File> R Script` and type the following code into the script editor of your RStudio session:

```{r EX1, echo=TRUE}
# this is a section header: 4 or more trailing dashes ---- 
# this is a comment: three or fewer trailing dashes --- 
# define a variable named patient 
patient <- 1  
# define a variable named ae 
ae <- "HEADACHE"  
# create a data frame from ToothGrowth 
tg <- ToothGrowth  
# Examine the first record 
head(tg)  
# Examine the last record 
tail(tg)  
# Examine a summary of the data frame tg 
summary(tg)
```

3.  Save your R script to the course directory (1.0 Getting Started). Name it RStudioTour.R

4.  Open the document outline using the `Show document` outline button. Click on the sections in the outline to navigate between them.

5.  Place your cursor on line 3 of your R script which is the comment above the patient variable definition. Insert a section by performing the following steps. From the RStudio menus, Go to `Code > Insert Section`. Type `Variable Definitions` into the dialogue box and click OK.

6.  Place your cursor on the line of code that reads: `patient <- 1` and click on the `Run` button in the upper right corner of your code editor.

7.  Check the Console Panel for any messages.

8.  Review the Environment Panel to view the value for the variable patient that was just created.

9.  In the upper right of the Environment Panel, switch from `List` view to `Grid` view and examine the information.

10. Check the box next to the patient variable and click on the `Clear objects from workspace` button with the broom icon.

11. Select all lines of the program and click the `Run` button.

12. Check the Console Panel for any messages.

13. Review the Environment Panel to view the value for the variables and data frames that were just created.

14. Place your cursor on the [`head()`](https://rdrr.io/r/utils/head.html) function in your program and hit the `F1` key to view the documentation on the head function.

15. Highlight the lines of code from `head(tg)` to `tail(tg)`. Go to `Code > Comment/Uncomment Lines`. Note the keyboard shortcut for this feature.

16. Save and close your script.
