---
title: "Using Shiny in R Markdown"
output: html_notebook
runtime: shiny
---

```{r info, include=FALSE}
# Packages & Functions on Display:
# - {shiny 1.6.0}:  selectInput, radioButtons, actionButton, HTML,
#                   reactive, eventReactive
```

<sub>
  Accel2R - Clinical R Training <br>
  © 2021 Experis ManpowerGroup Inc. All rights reserved
</sub>

### Getting Started

- Remember:
  - <i>You</i> can use <b>HTML</b> directly in the <code>rmarkdown</code> document
  - Use `markdown` shortcuts for common HTML styling like *italics* and **bold**
  - `Run` button to submit code to console
  - Help Menu > Markdown Quick Reference
  - Shortcut to insert code chunk: `Ctrl+Alt+I` / `Cmd+Option+I`

- Shiny
  - Notice `runtime: shiny` in RMD header
  - `Run Document` replaces `Knit`
  - R Markdown options control if output displays in Viewer or Pop-Up Window
  - Output is not "portable"
  - Document runs on a server (temporary/permanent)
  - R Markdown is 'busy' while document is running


```{r setup, include=FALSE}
# Set default RMarkdown options for entire document
knitr::opts_chunk$set(
  echo    = FALSE,     # Toggle R code display in final document
  message = FALSE,     # Toggle R function messages in final document
  warning = FALSE,     # Toggle R warnings in final document
  error   = FALSE)     # Toggle R errors in final document

# Load Packages
library(shiny)
library(tidyverse)
```

---

### Using Shiny Functions

- Shiny functions are generally just wrappers for HTML, JavaScript, and CSS
- Use `Run` to see the HTML code in R console
- Use `Run Document` to see the HTML application

```{r shiny-select, echo=TRUE}
shiny::selectInput(
  inputId = 'input_select',
  label   = 'Shiny: Select Something',
  choices = c('First Choice', 'Second Choice', 'Third Choice')
)
```

--- 

### Using Raw HTML

- Use HTML in `rmarkdown` text for a similar result

<label class="control-label" id="select_input-label" for="select_input">
  HTML: Select Something
</label>
<select id="select_input">
  <option value="First Choice" selected>First Choice</option>
  <option value="Second Choice">Second Choice</option>
  <option value="Third Choice">Third Choice</option>
</select>

### Using HTML Function

- Use `shiny::HTML()` to create customized HTML code using R / Shiny objects
- Use `paste` / `str_c` / `str_glue` to generate HTML code with R objects


```{r html-select}
shiny::HTML(
'<div>
  <label class="control-label" id="select_input-label" for="select_input">
    HTML: Select Something
  </label>
  <div>
    <select id="select_input">
      <option value="First Choice"  selected>First Choice  </option>
      <option value="Second Choice"         >Second Choice </option>
      <option value="Third Choice"          >Third Choice  </option>
    </select>
    <script type="application/json" data-for="select_input" data-nonempty="">
      {"plugins":["selectize-plugin-a11y"]}
    </script>
  </div>
</div>'
)
```


---

### Accessing User Input

- `inputId` must always be a unique identifier to store the user's selections
- Use the `input$` hidden global object to access various `inputId` values
- Wrap anything related to user input in `shiny::reactive()`

```{r shiny-input}
shiny::radioButtons(
  inputId = 'my_custom_radiobutton_name',
  label   = 'Select Something',
  choices = c('Apple', 'Banana', 'Cantelope')
)
```

You selected <b>`r shiny::reactive(input$my_custom_radiobutton_name)`</b>.

---

### What is `reactive`?

- `help(reactive, package = 'shiny')`
- "A reactive expression is an expression whose result will change over time"
- When a reactive object changes based on user input, all objects dependent upon that reactive object will change too

```{r shiny-reactive}
user_selection <- shiny::reactive(input$my_custom_radiobutton_name)

make_use_of_the_selection <- reactive(
  user_selection() %>% str_length()
)
```

You selected <b>`r user_selection`</b>, which has a string length of <b>`r make_use_of_the_selection`</b>.

- Note: 
  - In the above code I use `user_selection()` when inside an R code chunk and I use `user_selection` when including it in an 'inline' R operation. 
  - This is because `rmarkdown` assists with the interpretation of reactive objects
  - You won't be able to do this in a stand-alone `shiny` application


---

### Reactive Buttons

- Action buttons are initially set to `0` (for 0 times clicked)
- Values that depend on the button click are initially set to `NULL`

```{r shiny-button}

# Remember raw HTML can be used to customize this button
shiny::actionButton(
  inputId = 'action_button',
  label   = 'Button'
)

# Static Value
static_value <- 2

# Reactive Value
action_value <- eventReactive(
  eventExpr = input$action_button, 
  valueExpr = {static_value ^ input$action_button})
```

The initial value is `r static_value`, the button has been pressed `r reactive(input$action_button)` times.  The equation is `r static_value`<sup>`r reactive(input$action_button)`</sup> = `r action_value`.


### Examples of Other Output

```{r shiny-output}
shiny::selectInput(
  inputId  = 'select_var',
  label    = 'Select A Variable',
  choices  = colnames(mtcars),
  multiple = FALSE
)

# Subset Data
# Tidyverse Subset: reactive(mtcars %>% select(one_of(input$select_var)))
df_selection <- reactive(mtcars[[input$select_var]])
```

---

#### Raw R Output

```{r shiny-output-raw}
reactive(df_selection() %>% summary())
```

---

#### HTML Output

`r reactive(df_selection() %>% summary())`

---

#### Plot Output: 

`r reactive(df_selection() %>% hist())`

- **Note**: Anything that can't be used in `base::cat()` function needs special attention (dataframes, lists, plots, etc.)
- These cases will be covered in next lesson
  
  
---

### References

**Shiny Input Functions**

- Numeric Input
  - `numericInput()`
  - `sliderInput()`
- Date Input
  - `dateInput()`
  - `daterangeInput()`
- Text Input
  - `textInput()`
  - `textAreaInput()`
  - `passwordInput()`
- Categorical Input
  - `selectInput()`
  - `radioButtons()`
  - `checkboxInput()`
  - `selectizeInput()`
  - `varSelectInput()`
- Misc Input
  - `actionButton()`
  - `fileInput()`

---  

**Functions**

- `help(HTML,          package = 'shiny')`
- `help(reactive,      package = 'shiny')`
- `help(eventReactive, package = 'shiny')`
- `help(actionButton,  package = 'shiny')`
- `help(radioButtons,  package = 'shiny')`

---

**Website Links**

- [Site: Using Shiny in RMarkdown](http://rmarkdown.rstudio.com/authoring_shiny.html)
- [Site: RStudio Shiny Tutorial](https://shiny.rstudio.com/tutorial/)
- [Site: RStudio Shiny Cheatsheet](https://rstudio.com/resources/cheatsheets/)
- [Site: RStudio Shiny Reference](https://shiny.rstudio.com/reference/shiny/1.6.0/)
- [Book: Mastering Shiny](https://mastering-shiny.org/)
