---
title: "Getting Started with R"
author: "Anova Groups"
date: "2024-08-19"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# **2 R Language Foundation - Part I**

## **2.1 Introduction**

This course will focus specifically on Base R, using a bare minimum of external R packages. The content of this course is laid out in a building block approach, using topics consistently as we introduce them, and steadily increasing in complexity as we move forward.

When appropriate, we will provide side by side comparisons between SAS and the concepts we’re introducing in R.

In RStudio, we begin this course by emphasizing how to create RStudio projects to help organize our files, directories, and our RStudio working directory. The `working directory` is integral in managing how your scripts reference files stored on your filesystem.

Finally, we walk through the process of using the RStudio script editor and printing results to the console.

## **2.2 Names, Assignments, and Vectors**

This lesson covers the basics of naming rules in R, along with object assignment, data types, and vectors.

#### **Naming Conventions in R**

-   Names cannot begin with an underscore or a number

-   Names do accept letters, numbers, underscores, and periods

-   Names are case sensitive

-   The backtick (\`) can be used to create variable names that contain a space

#### **Assignments**

-   The [`assign()`](https://rdrr.io/r/base/assign.html) function can be used to assign values to an object

-   The `<-` operator is used as a shortcut for the [`assign()`](https://rdrr.io/r/base/assign.html) function

-   The `=` operator is only used for assignment inside of function arguments

#### **Data Types**

|                           |                  |
|---------------------------|------------------|
|                           | Data Value Types |
| Value                     | Type             |
| `1.2`                     | Numeric          |
| `1`                       | Numeric          |
| `1L`                      | Integer          |
| `"Hello"`                 | Character        |
| `"2020-01-01"`            | Character        |
| `as.Date("2020-01-01")`   | Date             |
| `TRUE`, `FALSE`, `T`, `F` | Logical          |
| `3 + 2i`                  | Complex          |
| `charToRaw("Hello")`      | Raw              |

-   Note that dates in R use the `YYYY-MM-DD` format by default.

#### **Vectors**

A vector is a set of values of the same datatype. We create a vector using the [`c()`](https://rdrr.io/r/base/c.html) command.

```{r}
v1 <- c(1, 2, 3) 
v2 <- c("one", "two", "three", "four")
```

We can check the class of a vector using [`class()`](https://rdrr.io/r/base/class.html) and [`typeof()`](https://rdrr.io/r/base/typeof.html), and we can check the length of a vector using [`length()`](https://rdrr.io/r/base/length.html).

```{r}
class(v1) 
typeof(v2) 
length(v2)
```

## **2.3 Operators**

#### **R Operators**

In this lesson we cover the various operators in R, including

-   Arithmetic Operators

-   Relational Operators

-   Logical Operators

-   Miscellaneous Operators

|                          |                                       |
|--------------------------|---------------------------------------|
|                          | Operators in R                        |
| Operator                 | Meaning                               |
| `+`, `-`, `*`, `/`       | Basic Arithmetic                      |
| `^`                      | Exponentiation                        |
| `%/%`                    | Quotient                              |
| `%%`                     | Modulo / Remainder                    |
| `==`, `!=`               | Equality                              |
| `>`, `>=`, `<`, `<=`,    | Greater Than, Less Than, and Equal To |
| `!`, `&`, `&&`, \|, \|\| | Logical Operators                     |
| `:`                      | Creating a Series (`1:10`)            |
| `%in%`                   | Check if Vector Contains an Element   |

#### **Vector Operations**

Nearly all functions in R are ‘vectorized’, meaning operations will be applied to each element of a vector.

```{r}
x <- c(2, 5, 4, 3) 
y <- c(1, 2, 4, 6)
```

Adding `x + y` will add each value to its corresponding element.

```{r}
x + y 
## [1] 3 7 8 9
```

Or, if we are using a single element object then its value is repeated.

```{r}
x + 100 
## [1] 102 105 104 103
```

## **2.4 Vectors**

In this lesson we cover vectors in detail. We begin by creating a numeric vector representing a collection of age values.

```{r}
age <- c(41, 53, 43, 39, 47, 52, 21, 38, 62, 26)
```

#### **Vector Subset**

We use the single bracket notation `[]` to subset vectors in R. For example, this code will return the 2nd element in the `age` vector.

```{r}
age[99] 
## [1] 53
```

And this code will remove the 2nd element of the `age` vector.

```{r}
age[-2] 
## [1] 41 43 39 47 52 21 38 62 26
```

#### **Vectorized Functions**

We cover various vectorized functions, including: `sort`, `sum`, `mean`, `median`, `min`, `max`, `range`, along with `names`, `length`, and `class`.

#### **Named Vectors**

```{r}
age_named <- c(s1 = 41, s2 = 53, s3 = 43, s4 = 39, s5  = 47, S6 = 52, S7 = 21, S8 = 38, S9 = 62, S10 = 26)
```

When appropriate, we can subset a vector using a name rather than an index. The result of these two commands is identical.

```{r}

age_named[2] 
age_named["s2"]
```

#### **Vector Decodes**

Referring to vector elements by name allows us to decode vectors using another mapping vector. In this case, we are converting the vector from `M` and `F` using the matching named vector to replace those values with `Male` and `Female`.

```{r}
# Create a Vector 
sex <- c("M", "F", "F", "M")  
# Create a Mapping Vector 
sex_decode <- c(M = "Male", F = "Female")  
# Apply the value mapping to existing vector 
sex_decode[sex]
##        M        F        F        M  
##   "Male" "Female" "Female"   "Male"
```

## **2.5 Factors and Lists**

In this lesson we introduce the data type `factors` and the data structure `lists`.

In R, **factors** are a special type of data structure used to represent **categorical data**. Categorical data refers to variables that have a fixed and limited set of values, often representing groups or categories. These categories can be either **unordered** (nominal) or **ordered** (ordinal).

#### **Simple Frequencies**

The best way to see how `factors` can be useful is by trying to get a frequency table for a categorical variable or vector. In the previous section, we introduced the concept of decoding values using a vector.

```{r}
arm <- c("A", "A", "A", "B", "A", "B", "A", "B", "A", "B") 
arm_decode <- c(A = "Placebo", B = "Active (50 mg)")  
arm_decode[arm]
```

We can use the [`table()`](https://rdrr.io/r/base/table.html) function to get counts for each value in the vector.

```{r}
table(arm_decode[arm]) 
## arm 
## A B  
## 6 4  table(arm_decode[arm]) 
##  
## Active (50 mg)        Placebo  
##              4              6
```

#### **Factor Category Levels**

But because these are just character values, values are counted as is. If we want to treat these values as levels of a category, where we can track zero counts of values that are not present in the data, this is when we want to use `factors`.

```{r}
arm_f <- factor(arm, levels = c("A", "B", "C", "D", "E"))  
table(arm_f) 
## arm_f 
## A B C D E  
## 6 4 0 0 0
```

#### Ordered Factors

In AE data, you may have ordinal categorical variables, like severity:

```{r}
trt01an <- factor(c("Placebo", "Drug A", "Placebo", "Drug B", "Drug A"), levels = c("Placebo", "Drug A", "Drug B", "Drug C"))
trt01an
aesev <- factor(c("None", "Mild", "Severe", "None", "Mild"), 
                        levels = c("None", "Mild", "Moderate", "Severe"), 
                        ordered = TRUE)
levels(aesev)

data <- data.frame(trt01an, aesev)

table(data$trt01an, data$aesev)


```

Using factor in graph

```{r}
trt01an <- factor(c("Placebo", "Drug A", "Drug A", "Placebo", "Drug B", "Drug B", "Placebo", "Drug A", "Drug B", "Placebo"), levels = c("Placebo", "Drug A", "Drug B", "Drug C"))
barplot(table(trt01an))
```

#### **Lists**

Vectors are used to store values of a single type. Lists can be used to store many vectors of varying types, using the [`list()`](https://rdrr.io/r/base/list.html) function.

**Unnamed List of Single Element Vectors**

```{r}
a_list <- list("a", 1, TRUE, as.Date("2020-01-01"))
a_list
```

**Named List of Vectors**

```{r}
b_list <- list(   letters = c("a", "b", "c"),   
                  numbers = c(1, 2, 3),   
                  dates   = c(as.Date("2020-01-01"), as.Date("2021-01-01")) )  
b_list 
## $letters 
## [1] "a" "b" "c" 
##  
## $numbers 
## [1] 1 2 3 
##  
## $dates 
## [1] "2020-01-01" "2021-01-01"
```

#### **Sub-setting a List**

**Key Differences:**

| Operation             | `[` (Single Bracket)                  | `[[` (Double Bracket)                                               |
|-----------------|--------------------|------------------------------------|
| **Returns**           | Sub-list (still a list)               | Actual object (contents of the list element)                        |
| **Multiple elements** | Can return multiple elements (a list) | Can only return a single element                                    |
| **Type returned**     | List                                  | The type of the object in the list (e.g., numeric, character, etc.) |
| **Indexing by name**  | Yes                                   | Yes                                                                 |
| **Example**           | `my_list[1:2]`                        | `my_list[[1]]`                                                      |

```{r}
# Create a list
my_list <- list(a = 1:3, b = c("X", "Y", "Z"), c = 42L)

names(my_list)
# Extract a sub-list using single bracket ([])
sub_list <- my_list[1]  # Returns a list containing the element `a`
sub_list
class(sub_list)
print(sub_list)

# Extract the actual contents of `a` using double bracket ([[)
element_a <- my_list[[1]]
print(element_a)
class(element_a)

# Instead of [[ ]] we can use $ (dollar sign) to fetch actual content
element_a_dollar <- my_list$a
class(element_a_dollar)
```

## **2.6 Data Frames**

In this lesson we introduce `data.frames`, data frame attributes, and subsetting data frames, along with giving an introduction to installing packages in R.

#### **Creating a Data Frame**

A data frame is similar to a list in that it is a collection of vectors. We begin by creating some vectors.

```{r}
library(randomNames)  
names  <- randomNames(4) 
subjid <- 101:104 
sex    <- factor(c("M", "F", "F", "M")) 
age    <- c(41, 53, 43, 39) 
arm    <- c(rep("A", 2), rep("B", 2))
```

Then, we use the [`data.frame()`](https://rdrr.io/r/base/data.frame.html) function to bring all of these vectors together.

```{r}
df <- data.frame(names, subjid, sex, age, arm)  
df 
##               names subjid sex age arm 
## 1   el-Elbaz, Noori    101   M  41   A 
## 2        James, Kia    102   F  53   A 
## 3    Moore, Cameron    103   F  43   B 
## 4 Johnston, Felisha    104   M  39   B
```

#### **Data Frame Attributes**

We can view the attributes of a data frame with various functions, some of the more common attribute accessors are show here:

```{r}
class(df) 
nrow(df) 
ncol(df) 
colnames(df) 
attributes(df)
```

#### **Data Frame Labels**

R does not have native support for variable labels like SAS does. However, there is a hidden `label` attribute within the data frame structure that we can access with the [`attr()`](https://rdrr.io/r/base/attr.html) function.

```{r}
# Modify the label attribute with the assignment operator 
attr(df$names, "label") <- "Subject Name"

view(df)
```

You can then view the labels using the `view()` function.

```{r}
# Instead of View(df), use one of the following options:

# Option 1: Display the first few rows with head()
head(df)
tail(df)
# Option 2: Print the entire data frame (useful for small data frames)
print(df)

# Option 3: Use knitr::kable() for a nicely formatted table
knitr::kable(df)
```

#### **Subsetting a Data Frame**

Just like with lists, we can use `$` to subset a data frame by variable name and return the chosen vector.

```{r}
df$names
```

We can also use the single brackets `[]` and double brackets `[[]]` to subset a data frame.

```{r}
df[1]; 
df["names"]  
df[[1]]; 
df[["names"]]
```

It is worth noting that the single bracket subset will return a data frame, while the double bracket subset will return a vector.

#### **Two Dimensional Subsetting**

We can subset a data frame by both columns and rows. In this case, we are requesting row 2 and the columns `names` and `sex`.

```{r}
df[2, c("names", "sex")]
```

We can also use the [`subset()`](https://rdrr.io/r/base/subset.html) function to subset a dataframe based on a logical condition.

```{r}
subset(df, select = c("names", "sex")) 
subset(df, age <=  40)
```

## **2.7 Capstone**

In this lesson, we walk through the process of creating a demographics dataframe using all of the concepts covered in this course.
