---
title: "R & SAS Comparison"
author: "Experis Accel2R - Clinical R Training"
output: 
  html_document:
    theme: united
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

# Packages
library(readr)
library(dplyr)
library(broom)
library(janitor)
library(magrittr)

# Data
# - https://bmcmedinformdecismak.biomedcentral.com/articles/10.1186/s12911-020-1023-5
# - Davide Chicco, Giuseppe Jurman: Machine learning can predict survival of
#   patients with heart failure from serum creatinine and ejection fraction alone.
#   BMC Medical Informatics and Decision Making 20, 16 (2020)

data_heart <-
  read_csv("data/heart.csv",
           col_types = cols(
             age                      = col_double(),
             anaemia                  = col_factor(),
             creatinine_phosphokinase = col_double(),
             diabetes                 = col_factor(),
             ejection_fraction        = col_double(),
             high_blood_pressure      = col_factor(),
             platelets                = col_double(),
             serum_creatinine         = col_double(),
             serum_sodium             = col_double(),
             sex                      = col_factor(),
             smoking                  = col_factor(),
             time                     = col_double(),
             DEATH_EVENT              = col_factor()
           ))


```

## Chi-Square Tests

#### Two x Two Table
```{r}
data_heart %>% tabyl(sex, diabetes) %>% knitr::kable()
```

#### Chi-Square Table Diagnostics
```{r}
chisq.test(data_heart$sex, data_heart$diabetes) %>% 
  augment() %>% 
  arrange(desc(data_heart.sex), data_heart.diabetes) %>% 
  relocate(.expected, .after = .observed) %>% 
  relocate(.resid, .std.resid, .after = .expected) %>% 
  knitr::kable()
```

#### Chi-Square Test
```{r}
chisq.test(data_heart$sex, data_heart$diabetes, correct = FALSE) %>% tidy() %>% knitr::kable()
```

#### Chi-Square Test with Correction
```{r}
chisq.test(data_heart$sex, data_heart$diabetes, correct = TRUE) %>% tidy() %>% knitr::kable()
```

#### Fisher Test, Greater Than
```{r}
fisher.test(data_heart$sex, data_heart$diabetes, alternative = "greater") %>% tidy() %>% knitr::kable()
```

#### Fisher Test, Less Than
```{r}
fisher.test(data_heart$sex, data_heart$diabetes, alternative = "less") %>% tidy() %>% knitr::kable()
```

#### Fisher Test, Two Sided
```{r}
fisher.test(data_heart$sex, data_heart$diabetes, alternative = "two.sided") %>% tidy() %>% knitr::kable()
```

#### Chi-Square Goodness of Fit
```{r}
data_heart %>%
  count(sex, name = "actual") %>%
  mutate(expected = ifelse(sex == 0, 0.40, 0.60)) %$%
  chisq.test(actual, p = expected) %>% 
  tidy() %>% 
  knitr::kable()
```

## Correlation

#### Correlation Matrix
```{r}
data_heart %>% select(where(is.numeric)) %>% cor() %>% as_tibble(rownames = "var_name") %>% knitr::kable()
```

#### Pearson Correlation
```{r}
data_heart %$% cor.test(age, platelets, method = "pearson") %>% tidy()  %>% knitr::kable()
```

#### Spearman Correlation
```{r}
data_heart %$% cor.test(age, platelets, method = "spearman") %>% tidy() %>% knitr::kable()
```

## T-Test

#### One-Sample T-Test
```{r}
t.test(data_heart$platelets, mu = 260000)               %>% tidy() %>% knitr::kable()
```

#### Two Sample T-Test
```{r}
t.test(platelets ~ sex, data = data_heart, var = TRUE)  %>% tidy() %>% knitr::kable()
```

#### Two-Sample Welch Test
```{r}
t.test(platelets ~ sex, data = data_heart, var = FALSE) %>% tidy() %>% knitr::kable()
```

#### Shapiro Wilk Test for Normality
```{r}
shapiro.test(data_heart$platelets) %>% tidy() %>% knitr::kable()
```

## ANOVA

#### One-Way ANOVA
```{r}
aov(platelets ~ sex, data = data_heart) %>% tidy() %>% knitr::kable()
```

#### One-Way ANCOVA
```{r}
aov(platelets ~ sex + age, data = data_heart) %>% tidy() %>% knitr::kable()
```

#### Two-Way ANOVA
```{r}
aov(platelets ~ sex + anaemia, data = data_heart) %>% tidy() %>% knitr::kable()
```

#### Factorial ANOVA
```{r}
aov(platelets ~ sex * anaemia, data = data_heart) %>% tidy() %>% knitr::kable()
```

#### ANOVA Diagnostics
```{r}
aov(platelets ~ sex * anaemia, data = data_heart) %>% 
  augment() %>% 
  slice(1:10) %>% 
  select(platelets, anaemia, sex, .fitted, .resid, .hat, .cooksd, .std.resid, .sigma) %>% 
  knitr::kable()
```

## Regression

#### Ordinary Least Squares Regression
```{r}
lm(platelets ~ age, data = data_train) %>% glance() %>% knitr::kable()
lm(platelets ~ age, data = data_train) %>% tidy() %>% knitr::kable()
```

#### Multiple Linear Regression
```{r}
lm(platelets ~ age + serum_sodium, data = data_train) %>% glance() %>% knitr::kable()
lm(platelets ~ age + serum_sodium, data = data_train) %>% tidy() %>% knitr::kable()

lm(platelets ~ age + serum_sodium, data = data_train) %>% augment(newdata = data_valid) %>% head() %>% knitr::kable()

lm(platelets ~ age + serum_sodium, data = data_train) %>% augment(newdata = tibble(age = 50, serum_sodium = 130, platelets = 225000)
) %>% head() %>% knitr::kable()


```

#### Logistic Regression
```{r}
glm(relevel(DEATH_EVENT, ref = "0") ~ age + platelets + smoking, data = data_train, family = "binomial") %>% glance() %>% knitr::kable()

glm(relevel(DEATH_EVENT, ref = "0") ~ age + platelets + smoking, data = data_train, family = "binomial") %>% tidy() %>% knitr::kable()

glm(relevel(DEATH_EVENT, ref = "0") ~ age + platelets + smoking, data = data_train, family = "binomial") %>% tidy(conf.int = TRUE, exponentiate = TRUE) %>% knitr::kable()

glm(relevel(DEATH_EVENT, ref = "0") ~ age + platelets + smoking, data = data_train, family = "binomial") %>% augment(new_data = data_valid, type.predict = "response") %>% 
  head() %>% knitr::kable()
```

