
<h1> <b>Using HTML tags in R Markdown<b> </h1>
<br>
<h1> Title1 Using HTML </h1>
<h2> Title2 Using HTML </h2>
<h3> <i>Title3 Using HTML</i> </h3>

---

# **Using R Markdown Instead**
# Title1 Using R Markdown
## Title2 Using R Markdown
### *Title3 Using R Markdown*

---

### Using RMarkdown to create nice bulleted lists

- Bullet 1
  - Sub-bullet 1
  - Sub-bullet 2

- Bullet 2
  - Sub-bullet 1

### Now let's use R and R Markdown together
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(DT)
bmi <- readRDS("../data/BMI_analysis.rds") %>% 
  select(SUBJECT, HEIGHT, WEIGHT)
```

```{r cars}
summary(bmi)
plot(bmi)
```

### Displaying Raw Results

```{r raw-output, echo = TRUE}
bmi %>% head(5)
```

### Displaying HTML Results

#### HTML Table - Using Kable
```{r kable-output, echo = TRUE}
# Run in RStudio to see HTML / Markdown table
bmi %>% head(5) %>% knitr::kable()
```

---

#### HTML Table - Using DT
```{r DT-output, echo = TRUE}
bmi %>% DT::datatable()
```


