---
title: "Using HTML in R Markdown"
output: html_document
editor_options: 
  chunk_output_type: inline
---


### R Markdown Tour

- Create R Markdown Document
  - File > New File > R Markdown
  - Create R Markdown with Shiny
  - Create R Markdown with Template

- Execution
  - `Knit` creates the document
  - `Run` sends R code to the console
  - Output is portable HTML, DOCX, PDF, RTF file
  
- Knit Options
    - Knit Directory controls the working directory when running a document
    
- R Markdown Options
  - Preview in RStudio Viewer or in pop-up window
  - Show output either in console or under the code itself
  - Output options
    - Select output format
    - Toggle table of contents, numbered headers
    - Custom themes, custom CSS themes
    - Default figure output controls
    
- Learn More
  - RStudio Help > Markdown Quick Reference
  
---

<sub>
  Experis Accel2R - Clinical R Training<br>
  © 2021 Experis ManpowerGroup Inc. All rights reserved
</sub>


<h1>Using HTML</h1>
  - <h2>Here's an HTML level 2 subheader</h2>
    - <h3>Here's an HTML level 3 subheader</h3>
      - <p>This is my paragraph text with <b>bold</b> and <i>italics</i>.</p>
<hr>

# Using Markdown Shortcuts

No need to include HTML tags `rmarkdown` takes care of that in the background.

- Use code formatting using `backticks`
- Headers with `#`
- Bullet lists with `-`
- *Italics* with `*`
- **Bold** with `**`
- Horizontal rule with `---`

---

# Header Level 1
## Header Level 2
### Header Level 3

---

### Using R Inside RMarkdown

- Adding code chunks
  - Use `+C` Menu (R, Python, etc.)
  - Using shortcut `Ctrl+Alt+I` / `Cmd+Option+I`

- Note:
  - Interactive coding environment is separate from RMD execution environment
  - Every R object or package required for the final output needs to be included in the document.

```{r setup, include=FALSE}
# Always provide label for code chunk

# include=TRUE  : code and output will be included in final document
# include=FALSE : code and output will not be included in final document

# Set default options for entire document
knitr::opts_chunk$set(
  echo    = FALSE,     # Toggle R code display in final document
  message = FALSE,     # Toggle R function messages in final document
  warning = FALSE,     # Toggle R warnings in final document
  error   = FALSE)     # Toggle R errors in final document

# Load packages in setup
library(tidyverse)
library(plotly)        # For interactive graphs
library(DT)            # For interactive tables
```

### Displaying Raw Results

```{r raw-output, echo = TRUE}
mtcars %>% head(5)
```

### Displaying HTML Results

#### HTML Table - Using Kable
```{r kable-output, echo = TRUE}
# Run in RStudio to see HTML / Markdown table
mtcars %>% head(5) %>% knitr::kable()
```

---

#### HTML Table - Using DT
```{r DT-output, echo = TRUE}
mtcars %>% DT::datatable()
```

---

### Using R Inline

- We can use `backticks` to run R code in the written text
- This allows static reports to be updated dynamically when the data is updated
- For debugging purposes, test this code before 'knitting' the document

For example, in the dataframe **mtcars** there are `r nrow(mtcars)` rows.  The columns in this dataframe are <i>`r mtcars %>% colnames() %>% str_c(collapse = ', ')`</i>.  The mean *mpg* is <b>`r mean(mtcars$mpg) %>% round(2)`</b>.


### Using Math Expressions

- Use the dollar sign to include math notation
  - Bars:          $1.\overline{3}$
  - Fractions:     $\frac{1}{2}$
  - Superscript:   $x^2$
  - Greek Letters: $\chi \lambda \beta$
- Learn more about math expressions in RMD  [here](https://rpruim.github.io/s341/S19/from-class/MathinRmd.html)

---

### Displaying Other Results

#### Static Plots using `ggplot2`
```{r ggplot}
mtcars %>% ggplot(aes(x = mpg, y = disp)) + geom_point()
```

#### Interactive Plots using `plotly`
```{r plotly}
mtcars %>% plot_ly(x = ~mpg, y = ~disp) %>% add_markers()
```


### RMarkdown via Function

- Create documents within another script, function, or loop
- Use `params` to specify parameters and arguments to pass into document generation function

```{r, echo = TRUE}
# rmarkdown::render('path/to/document.rmd', params = )
```

### Resources and Documentation

- [W3 Schools HTML Reference](https://www.w3schools.com/html)
- [R Markdown at the RStudio Website](http://rmarkdown.rstudio.com)
- [R Markdown Free Online Book](https://bookdown.org/yihui/rmarkdown/)
