# Quick Regular Expressions (Regex) Tutorial in R

Regular expressions (regex) are powerful tools for pattern matching and text manipulation. In R, they are extensively used for searching, extracting, and replacing text. This tutorial provides a concise guide to using regex in R.

## 1. Basic Syntax

### Literal Characters

-   **Literal characters** match themselves.

```{r}
aedecod <- c("Neck Pain", "Stomach Pain", "Heacache")
grep("pain", aedecod, ignore.case = TRUE)  # Matches "pain" in all words

```

### Metacharacters

-   **Metacharacters** have special meanings

```{r}
# . ^ $ * + ? { } [ ] \ | ( )
files <- c("file.txt", "data.csv")
grep("\\.", files)  # Matches strings containing a dot
grep(r"(.)",files)
```

## 2. Special Characters

### Dot `.`

-   Matches **any single character** except a newline

    ```{r}
    grep("c.t", c("cat", "cut", "coat"))  # Matches "cat" and "cut"

    ```

### Anchors `^` and `$`

-   `^` matches the **start** of a string.

-   `$` matches the **end** of a string

    ```{r}
    grep("^cat", c("cat", "concatenate"))  # Matches "cat" at the beginning
    grep("cat$", c("scat", "concat"))      # Matches "concat" ending with "cat"

    ```

## 3. Character Classes

### Square Brackets `[ ]`

-   Match any **one character** inside the brackets

    ```{r}
    grep("b[aeiou]t", c("bat", "bet", "bit", "but"))  # Matches all listed words

    ```

### Negation `[^ ]`

-   `[^ ]` matches any character **not** inside the brackets

```{r}
grep("b[^a]t", c("bat", "bet", "bit"))  # Matches "bet" and "bit"

```

### Predefined Classes

-   `\\d` : Digit `[0-9]`

-   `\\D` : Non-digit `[^0-9]`

-   `\\w` : Word character `[a-zA-Z0-9_]`

-   `\\W` : Non-word character `[^a-zA-Z0-9_]`

-   `\\s` : Whitespace `[ \t\n\r\f\v]`

-   `\\S` : Non-whitespace `[^ \t\n\r\f\v]`

**Note:** In R, backslashes need to be escaped, so `\d` becomes `\\d`.

```{r}
grep("\\d+", c("item1", "item2", "item"))  # Matches "item1" and "item2"

```

## 4. Quantifiers

### Symbols

-   `*` : **0 or more** occurrences

-   `+` : **1 or more** occurrences

-   `?` : **0 or 1** occurrence

-   `{n}` : **Exactly n** occurrences

-   `{n,}` : **n or more** occurrences

-   `{n,m}` : **Between n and m** occurrences

```{r}
grep("go+", c("go", "goo", "good"))  # Matches "go" and "goo"
grep("colou?r", c("color", "colour"))  # Matches both spellings


```

## 5. Grouping and Alternation

### Parentheses `( )`

-   **Group** parts of a pattern

    ```{r}
    # Install and load stringr package if not already installed
    # install.packages("stringr")
    library(stringr)

    # Example data
    texts <- c('Cycle 11, Day 1', 'Cycle 12, Day 2')

    # Regular expression to match and capture cycle and day numbers
    pattern <- "Cycle (\\d+), Day (\\d+)"


    # Extract matches (including capture groups)
    matches <- str_match(texts, pattern)

    # View the result
    matches


    cycle <- matches[,2]
    cycle

    days <- matches[,3]
    days
    ```

### Pipe `|`

-   **Alternation**, matches either the left or right expression

    ```{r}
    grep("cat|dog", c("cat", "dog", "bird"))  # Matches "cat" and "dog"
    grepl("cat|dog", c("cat", "dog", "bird"))
    ```

## 6. Lookaround (Advanced)

-   **Lookahead** `(?=...)`

-   **Negative Lookahead** `(?!...)`

-   **Lookbehind** `(?<=...)`

-   **Negative Lookbehind** `(?<!...)`

> **`Note:`**`Lookaround assertions may not be fully supported in all R regex functions.`

## 7. Common R Functions Using Regex

| Function     | Use                                                |
|--------------|----------------------------------------------------|
| `grep()`     | Searches for matches and returns the **indices**   |
| `grepl()`    | Returns a **logical vector** indicating matches.   |
| `sub()`      | replaces the **first** match.                      |
| `gsub()`     | replaces **all** matches                           |
| `regexpr()`  | Returns the **position** and **length** of matches |
| `strsplit()` | Splits strings based on a regex pattern            |

```{r}
regexpr("cat|dog", c("cat", "dog", "bird"))
gregexpr("cat|dog", c("cat", "dog", "bird"))
```
